---
title: "PA1_template"
author: "Jonathan Isernhagen"
date: "October 13, 2015"
output: html_document
---

This document details an investigation into the exercise patterns of a device-wearing volunteer.

Step 1:  set up: the "activity.csv" file must be loaded in working directory in order for this to function, and the dplyr package must be loaded.

```{r}
require(dplyr)
activity<- read.csv("activity.csv")
activity$interval<-as.factor(activity$interval)
```

Step 2a):  Calculate the total steps taken per day
```{r}
by_date <- group_by(activity, date)
activitydaysum <- summarise(by_date, count = n(), stepsum = sum(steps, na.rm = TRUE), stepmean = mean(steps, na.rm = TRUE), stepmedian = median(steps, na.rm = TRUE))
activitydaysum
```

Step 2b): create a histogram plot of the mean and median of the total number of steps taken each day
```{r}
par(mfrow = c(1, 2))
with(activitydaysum, {
  hist(stepmean, col="red", xlab = "Mean", main = "Mean of Non-null Steps")
  hist(stepmedian, col="red", xlab = "Median", xlim = c(-2, 2), main = "Median of Non-null Steps")
})
```
It is apparent from this plot that there are many intervals of complete inactivity, rendering the median almost useless.

Step 3a) Make a time series plot of 5-minute interval (on the x-axis) and avg number of steps taken, averaged across all days (on the y-axis)
```{r} 
by_interval <- group_by(activity, interval)
activityintervalsum <- summarise(by_interval, count = n(), stepsum = sum(steps, na.rm = TRUE), stepmean = mean(steps, na.rm = TRUE), stepmax = max(steps, na.rm = TRUE))
```

3b) Calculate which 5-minute interval, on average across all days in the dataset, contains the maximum number of steps?
```{r}
activityintervalsum<-arrange(activityintervalsum, desc(stepmean))
slice(activityintervalsum, 1)
```

4a) Calculate and report the total number of missing values in the dataset
```{r}
sum(is.na(activity$steps))
```

4b) Explain strategy for imputing missing values:  time-of-day and day-of-week effects seem highly pronounced, so the strategy was to replace nulls with the mean for each null value's weekday and time.

4c) Create new dataframe using this strategy to fill in missing values.
```{r}
activityfilled <- activity #1) create data frame to fill with imputed step values
activityfilled <- within(activityfilled, date <- as.POSIXct(paste(date))) #2) change date to POXIXct format 
activityfilled <- within(activityfilled, weekday <- weekdays(as.Date(date))) #3) add a day of week variable
activityfilled <- within(activityfilled, weekend <- "weekday") #4) add a weekday/weekend variable, defaulted to weekday
activityfilled$weekend[activityfilled$weekday == "Saturday"] <- "weekend" #5) overwrite "weekend" for Saturdays
activityfilled$weekend[activityfilled$weekday == "Sunday"] <- "weekend" #6) overwrite "weekend" for Sundays
activityfilled <- within(activityfilled, stepsfilled <- steps) #7) fill steps values into the "stepsfilled" column
by_day_interval <- group_by(activityfilled, weekday, interval) #8) create filler table:  group by weekday and interval
filler <- summarise(by_day_interval, stepsfill = mean(steps, na.rm = TRUE)) #9) create filler table:  add mean(steps) by weekday and interval
activityfilled <- inner_join(activityfilled, filler, b = c("weekday","interval")) #10) join data frames
activityfilled$stepsfilled[is.na(activityfilled$stepsfilled)] <- activityfilled$stepsfill[is.na(activityfilled$stepsfilled)] #11) write weekday+interval mean values over nulls
```

4d) Calculate/report the mean and median total steps taken each day.  Calculate/report difference vs. unimputed values
```{r}
by_date_new <- group_by(activityfilled, date)
activitydaysumnew <- summarise(by_date_new, count = n(), stepsum = sum(stepsfilled, na.rm = TRUE), stepmean = mean(stepsfilled, na.rm = TRUE), stepmedian = median(stepsfilled, na.rm = TRUE))
activitydaysumnew
```

4e) Plot histogram of steps taken each day
```{r}
par(mfrow = c(1, 2))
with(activitydaysumnew, {
  hist(stepmean, col="red", xlab = "Mean", main = "Mean of Non-null Steps")
  hist(stepmedian, col="red", xlab = "Median", xlim = c(-2, 2), main = "Median of Non-null Steps")
})
```

5a) Add a weekend/weekday factor variable to imputed data set (done upstream).

5b) Make a panel plot of time series (type = 1) plot with 5-minute interval on X-axis, avg. steps per weekend/weekday day on y.
```{r}
weekend<-filter(activityfilled, weekend == "weekend")
weekday<-filter(activityfilled, weekend == "weekday")
by_interval_weekend <- group_by(weekend, interval, weekend)
activityintervalsumday <- summarise(by_interval_weekend, count = n(), stepsum = sum(steps, na.rm = TRUE), stepmean = mean(steps, na.rm = TRUE), stepmax = max(steps, na.rm = TRUE))
by_interval_weekday <- group_by(weekday, interval, weekend)
activityintervalsumend <- summarise(by_interval_weekday, count = n(), stepsum = sum(steps, na.rm = TRUE), stepmean = mean(steps, na.rm = TRUE), stepmax = max(steps, na.rm = TRUE))
par(mfrow = c(1, 2))
  plot(activityintervalsumday$stepmean, xlab = "Intervals in Day", ylab = "Mean steps", main = "Weekday")
  plot(activityintervalsumend$stepmean, xlab = "Intervals in Day", ylab = "Mean steps", main = "Weekend")
```

